---
title: "Creating Hexagonal Grid Representations of Irregular Polygons"
author: "Johan Junkka"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating Hexagonal Grid Representations of Irregular Polygons}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## The Challenge: Visualizing Data with Uneven Spatial Units

When working with administrative boundaries like parishes, municipalities, or census tracts, we often encounter a common problem: the spatial units vary drastically in size. In Sweden, for example, parishes can range from tiny urban units like Stockholm domkyrka to vast rural parishes like Jukkasjärvi. This size disparity creates challenges when visualizing data:

1. Small units become nearly invisible on maps
2. Large units dominate the visual space
3. The actual geographical patterns can be hard to discern

This vignette demonstrates how to transform such irregular spatial units into a more standardized hexagonal grid representation while maintaining spatial relationships.

## The Solution: Hexagonal Grid Transformation

Our approach converts irregular polygons into a hexagonal grid representation where:

- Small polygons are combined into single hexagons
- Large polygons can span multiple hexagons
- Spatial relationships are preserved
- The visualization becomes more standardized

Let's walk through the development process.

## Loading Required Libraries

```{r libraries}
library(sf)
library(histmaps)
library(tidyverse)
```

## Step 1: Creating the Base Hexagonal Grid

First, we need to create a hexagonal grid that covers our study area. The size of the hexagons is a key parameter that affects the final representation:

```{r hex-grid}
# Set the hex size (in the same units as your projection)
size <- 25000/1.2  # Approximately 20.8 km

# Create base area from the county boundaries
counties <- geom_sp %>% 
  filter(start <= 1900, end >= 1900, type_id == "county") %>% 
  summarise(1)

# Create hex grid
hex_grid <- st_make_grid(counties, 
                        square = FALSE, 
                        cellsize = size, 
                        what = "polygons")

# Plot to examine
plot(st_geometry(hex_grid), main = "Initial Hex Grid")
plot(st_geometry(counties), add = TRUE, border = "red")
```

## Step 2: Cleaning the Grid

The initial grid includes many hexagons that only partially overlap with our study area. We want to keep only hexagons that have meaningful coverage:

```{r clean-grid}
# Get intersections
hex_sf <- hex_grid[counties] %>%
  st_as_sf() %>%
  rownames_to_column(var = "id")

base_area <- st_area(hex_sf$x[1]) %>% as.numeric()

# Calculate coverage and filter
res0 <- st_intersection(hex_sf, counties)
hex_coverage <- res0 %>%
  group_by(id) %>%
  summarise(
    n = n(),
    area = sum(as.numeric(st_area(.)))
  )

# Keep hexes with >20% coverage
valid_hexes <- hex_coverage %>%
  filter(area > (base_area * 0.2))

hex_clean <- hex_sf %>% 
  filter(id %in% valid_hexes$id)

plot(st_geometry(hex_clean), 
     main = "Cleaned Hex Grid (>20% Coverage)",
     col = "#ff000088")
```

## Step 3: Assigning Parishes to Hexagons

The core challenge is determining how to assign parishes to hexagons. We identified several cases that needed different handling:

1. One-to-one relationships: Parishes that fit within a single hex
2. Many-to-one: Multiple small parishes in one hex
3. One-to-many: Large parishes spanning multiple hexes
4. Edge cases: Parishes with partial coverage

Here's how we handle each case:

```{r parish-assignment}
# Get parish data for 1930
parishes <- geom_sp %>% 
  filter(type_id == "parish", start <= 1930, end >= 1930)

# Calculate intersections
intersections <- st_intersection(parishes, hex_clean) %>%
  mutate(
    area = as.numeric(st_area(.)),
    full_coverage = area >= base_area
  ) %>%
  rownames_to_column() %>%
  group_by(geom_id) %>%
  mutate(
    n_hexes = n(),
    n_full = sum(full_coverage)
  ) %>%
  ungroup()

# 1. Handle one-hex parishes
one_hex <- intersections %>% 
  filter(n_hexes == 1)

# 2. Handle multi-hex parishes with no full coverage
other_hex <- intersections %>% 
  filter(n_hexes > 1)

no_full_hex <- other_hex %>%
  arrange(-area) %>%
  group_by(geom_id) %>%
  filter(max(full_coverage) != 1) %>%
  slice(1) %>%
  ungroup()

# 3. Handle multi-hex parishes with full coverage
full_hex <- other_hex %>%
  group_by(id) %>%
  filter(n() == 1) %>%
  ungroup()
```

## Step 4: Handling Edge Cases

For parishes that don't clearly fit into our main categories, we use area-based assignment:

```{r edge-cases}
# Find remaining cases
other_hex2 <- other_hex %>%
  filter(!id %in% unique(full_hex$id)) %>%
  filter(!id %in% unique(no_full_hex$id)) %>%
  filter(!id %in% unique(one_hex$id))

# Assign based on largest area of intersection
leftover_hex <- other_hex2 %>%
  arrange(-area) %>%
  group_by(id) %>%
  slice(1) %>%
  ungroup()

# Combine all assignments
hex_data <- bind_rows(one_hex, full_hex, no_full_hex, leftover_hex)
```

## Step 5: Creating the Final Output

The final step is to create a clean data structure that links hexagons to their assigned parishes:

```{r final-output}
# Create linkage table
hex_links <- hex_data %>% 
  st_drop_geometry() %>%
  select(geom_id, id)

# Create hex summary
hex_summary <- hex_data %>%
  st_drop_geometry() %>%
  group_by(id) %>%
  summarise(
    geom_ids = paste(unique(geom_id), collapse = "-"),
    area_tot = sum(area)
  ) %>%
  mutate(
    prop = area_tot/base_area
  )

# Final hex spatial object
final_hex <- hex_clean %>%
  left_join(hex_summary, by = "id")

# Visualize final result
ggplot(final_hex) +
  geom_sf(aes(fill = prop)) +
  scale_fill_viridis_c() +
  theme_minimal() +
  ggtitle("Final Hexagonal Grid with Coverage Proportions")
```

## Conclusions

This approach provides several benefits:

1. More standardized visualization units
2. Better representation of both small and large parishes
3. Maintained spatial relationships
4. Flexible assignment rules based on coverage and area

The method can be adapted for different spatial scales by adjusting:

- Hex size
- Minimum coverage threshold
- Assignment rules for edge cases

## Using the Package

The entire process has been encapsulated in the `create_hex_map()` function:

```{r package-usage, eval=FALSE}
result <- create_hex_map(
  sf_object = parishes,
  hex_size = 25000,
  min_coverage = 0.2,
  id_col = "geom_id",
  name_col = "name"
)
```

This provides a reusable solution for similar spatial aggregation challenges.
